Anomaly detection with kernel density estimation on manifolds

### 2022-02-26 paper/paper.tex
5d plots and tables done; add colors as distance to center

For the high-dimensional case, we generate the meta data from a 5-dimensional semi-hypersphere. First, we simulate \(N=2000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(25,25,25,25)\). In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as 15. Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix. Matrix \(Q\) is used as the rotation matrix and it satisfies
\(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_isomap_riem20}

@@

\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
\centering
\begin{tabular}[t]{l>{}l>{}l>{}l}
\toprule
  & ISOMAP & LLE & UMAP\\
\midrule
Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{-0.130}\\
Fixed bandwidth & 0.806 & 0.940 & -0.341\\
\bottomrule
\end{tabular}
\end{table}

>>>>

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space. First, we simulate \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\). In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as \(7\). Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix. Matrix \(Q\) is used as the rotation matrix and it satisfies
\(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.

@@

{\centering \includegraphics[width=0.95\linewidth]{figures/fived_density_comparison_4ml_riem20}

@@

\caption{\label{tab:fivedcors}Correlation between true density and estimated density for four manifold learning embeddings.}
\centering
\begin{tabular}[t]{l>{}l>{}l>{}l>{}l}
\toprule
  & ISOMAP & LLE & Laplacian.Eigenmaps & UMAP\\
\midrule
Variable bandwidth & \textbf{0.921} & \textbf{0.981} & \textbf{0.662} & \textbf{-0.130}\\
Fixed bandwidth & 0.806 & 0.940 & -0.181 & -0.341\\
\bottomrule
\end{tabular}
\end{table}
